---
  title: "glitr: How to adorn your plots"
  description: This vignette covers the core features of the glitr package, focusing on how to add SIEI recommended colors to different types of plots generated by ggplot2.
  output: rmarkdown::html_vignette
  vignette: >
    %\VignetteIndexEntry{glitr: How to adorn your plots}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
### Introduction

This vignette provides best practices for applying SIEI recommended colors to visualizations using the `glitr` package. 

### Getting Started

To get started, load the standard OHA-SI libraries that are often used in analysis. 
```{r setup, echo = T, eval = F}
library(tidyverse)
library(glitr)
library(scales)
library(ggtext)
```

Take a look at what functions and objects are included with `glitr'.
```{r explore glitr, echo = T, eval = T}
ls('package:glitr')
```

The `glitr` package can be divided into three main parts i) colors, that come bundled as objects, ii) si themes, that can be used to quickly apply SI plot defaults, and iii) helper functions, that interpolate palettes or apply palettes to ggplots. We will explore the colors first.

### Colors
A number of pre-defined colors come with the `glitr` package. The set that is probably of most interest is the SIEI recommended colors. As you may recall from the [Data Visualization Style Guild](https://usaid-oha-si.github.io/styleguide/), the SI team has created seven core colors: [denim](https://via.placeholder.com/50/2057a7/000000?text=+), old_rose, moody blues, burnt sienna, scooter, golden sand and genoa. 
