---
  title: "glitr: How to adorn your plots"
  description: This vignette covers the core features of the glitr package, focusing on how to add SIEI recommended colors to different types of plots generated by ggplot2.
  output: rmarkdown::html_vignette
  vignette: >
    %\VignetteIndexEntry{glitr: How to adorn your plots}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  warning = FALSE, 
  message = FALSE)
```
### Introduction

This vignette provides best practices for applying SIEI recommended colors to visualizations using the `glitr` package. 

### Getting Started

To get started, load the standard OHA-SI libraries that are often used in analysis. 
```{r setup, echo = T, eval = T}
library(tidyverse)
library(glitr)
library(scales)
library(ggtext)
library(extrafont)
```


```{r what is inside glitr, echo = T, eval = T}
# Take a look at what's inside the package
ls('package:glitr')
```

The package can be divided into three main parts i) colors, that come as objects (such as grey30k or usaid_blue), ii) SI themes, that can be used to quickly apply SI plot defaults, and iii) helper functions, that interpolate palettes or apply palettes to a visualization. This vignette will focus on exploring colors and the SI themes. 

### Colors
A number of pre-defined colors come with the `glitr` package. All objects starting with 'grey' belong to a family of gray colors where `grey10k` is the lightest and `grey90k` the darkest. Objects starting with 'usaid_' are the official USAID colors while those starting with 'wapo_' are Washington Post inspired colors. `color_` flags predefined colors that can be used to for filling in captions, gridlines, plot text or plot titles.  

```{r}
# Colors belonging to greys, usaid_, wapo_ or siei_.
grep("(grey|siei_|wapo_|usaid_)", ls('package:glitr'), value = T)
```


The set that is probably of most interest is the SIEI recommended colors. As you may recall from the [Data Visualization Style Guild](https://usaid-oha-si.github.io/styleguide/), the SI team has created seven core colors:

![#2057a7](https://via.placeholder.com/10/2057a7/000000?text=+) denim  
![#c43d4d](https://via.placeholder.com/10/c43d4d/000000?text=+) old_rose  
![#8980cb](https://via.placeholder.com/10/8980cb/000000?text=+) moody_blue    
![#e07653](https://via.placeholder.com/10/e07653/000000?text=+) burnt_sienna  
![#1e87a5](https://via.placeholder.com/10/1e87a5/000000?text=+) scooter  
![#f2bc40](https://via.placeholder.com/10/f2bc40/000000?text=+) golden_sand   
![#287c6f](https://via.placeholder.com/10/287c6f/000000?text=+) genoa  

The hex codes for all these colors are added to your workspace when you load `glitr`. 
```{r show colors, echo = T, fig.height = 2, fig.width = 2}
# Any of these colors can be called by simply typing in the name of the color. 
# The `show_col` function is from the `scales` package and is a handy way to preview a color. 
show_col(genoa)
show_col(c(denim, old_rose, moody_blue, burnt_sienna), borders = F)
```

To access to full list of discrete or continuous palettes, see the si_palettes list or call the color palette directly using the name. Color names that are singular (genoa, old_rose) are categorical palettes based on suggested color pairs. Color names that are plural (genoa) are continuous palettes that can be applied to continuous variables. If you attempt to apply a discrete palette to a continuous variable, the color pairs will be recycled to the length of the vector you are attempting to encode.

```{r palettes, echo = TRUE, fig.height = 2, fig.width= 3}
si_palettes$genoa %>% show_col(labels = F, borders = F)
si_palettes$genoas %>% show_col(labels = F, borders = F)
```

Finally, if you want to create a custom palette from one of the color sets in the si_palettes list, you can do this using the `si_rampr()` function. 

```{r custom palettes, fig.height = 2, fig.width= 3}
# si_rampr takes a palette name and the number of interpolated colors (n) you wish to return as arguments. 
denim_pal <- si_rampr(pal_name = "denims", n = 25)
denim_pal

show_col(denim_pal, labels = FALSE, borders = FALSE)

```


### Applying Colors

To add any of these colors to a graph, pass them as arguments in a ggplot2 call. We will work with one of the sample data sets to create a bar graph where positive testing results are colored with scooter. 

```{r bar graph with colors, fig.height = 3, fig.width = 6}
# Munge the hts data down to testing yields for a given year
  hts_bar <-
    hts %>%
    filter(period == "FY50", period_type == "cumulative") %>%
    group_by(indicator, primepartner) %>%
    summarise(total = sum(value, na.rm = TRUE)) %>%
    ungroup() %>%
    spread(indicator, total) %>%
    mutate(yield = (HTS_TST_POS / HTS_TST),
           prime_order = fct_reorder(primepartner, HTS_TST)) %>% 
  arrange(prime_order)

# Define testing results to be grey30k and positive results to be scooter. 
# Play close attention to where the fill is placed in the geom_col() call. 
# If placed inside the aesthetics, you will need to apply scale_fill_identity() to get the colors to render.
  p <- hts_bar %>% 
    ggplot(aes(y = prime_order)) +
    geom_col(aes(x = HTS_TST), fill = grey30k) +
    geom_col(aes(x = HTS_TST_POS), fill = scooter) +
    labs(x = NULL, y = NULL, title = "HTS_TST_POS FILLED WITH SCOOTER COLOR") +
    coord_cartesian(expand = F) # Move names closer to y-axis
  p
```

### SI Themes
In order to create visualizations that appear to belong to the same family ([Think Baldwin Brothers](https://live.staticflickr.com/4111/5195907338_df50404c30_b.jpg)), `glitr` includes nine ggplot themes that simplify a plot down to its core elements. At the base of these themes is the `si_style`. 

```{r}
grep("(si_style)", ls('package:glitr'), value = T)
```


We can see how this differs from the default ggplot2 theme by applying the theme to the plot above.

```{r si_style, fig.height = 3, fig.width = 6}
p + si_style()
```

You will notice the si_style strips the background gray color and converts the grid lines to a light shade of gray. As this plot is oriented horizontally, we can remove the extra y gridlines by using the `si_style_xgrid` theme. All of the si themes also come with pre-formatted titles, subtitles, and captions following the style guild standards. 

```{r, fig.height = 3, fig.width = 6}
p + 
  si_style_xgrid() + 
  scale_x_continuous(labels = comma)+
  labs(caption = paste("Generated on ", Sys.Date(), " by the SI team."),
      subtitle = "Not real data, but it still looks pretty good.") 
```


