---
title: "What the f*nt?"
output: rmarkdown::html_vignette
description: |
  This vignette covers how to access Source Sans Pro (or other fonts) and use them in R
vignette: >
  %\VignetteIndexEntry{what-the-f_nt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>"
)
```
To create standard visualizations across our team, we rely on one of USAID's alternate fonts, Sans Source Pro. This font is not only not native to R, nor is it a standard to Windows. 

### Downloading/Installing
Sans Source 3 (formerly Pro) is an open source font available from Google Fonts and from Software Center on USAID GFEs. 
  - To install from Google Fonts, navigate to this [link](https://fonts.google.com/specimen/Source+Sans+3) and click the "Download family". After the folder finishes downloading, unzip it. 
  - To install on your GFE, open Softare Center App, click on "Source Sans Pro (Google Web Font)", and then select install.

### Accessing Fonts in R 
To use non-native fonts in R, you must run a program called `extrafonts` which is installed as a dependency from `glitr`. You will need to run the following code below to install all the fonts on your computer (if desired), including your newly downloaded Source Sans 3 (fomerly Pro) typeface. You will only need to import fonts only once. However, to use these fonts with any plotting in R, you will need to load the `extrafont` library as with any other package.

```{r}
# Required library
library(extrafont)
library(remotes)

# Downgrade a package dependency for extrafont
#https://stackoverflow.com/questions/61204259/how-can-i-resolve-the-no-font-name-issue-when-importing-fonts-into-r-using-ext/68642855#68642855
install_version("Rttf2pt1", version = "1.3.8")

#import all Windows fonts
  font_import()
  
#restart your R session - CTRL + SHIFT + F10

library(extrafont)
fonts()
```
